Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-14 demystify.tex
updates

\chapter{Letters of Recommendation (LOR)}\label{chap:lor}
\chapterinfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

Most PhD programs will require at least \textbf{two LORs}. 
The goal of the LoR is to provide the adcom with an assessment of your research ability and potential. 
The letter should 
focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher.

@@

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also \emph{have the credibility} to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD (so they know what a PhD is about) and is active in research (so they can evaluate your research ability). LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\subsection{LoR from Famous People}  Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school). The reason is simple: adcom members trust people that they know and respect. If a well-known researcher enthusiastically recommends you, then the adcom members will likely trust that recommendation.

@@

It is also worth noting that a LoR from a ``full'' senior professor who has not been active in research might not be as strong as a LoR from a junior faculty who has been actively mentoring you in research.  This also applies to administrative, e.g., chairs, directors, and deans (\autoref{sec:admin-letters}.) The fame of the writer is less important than how well they know you and can write about your research ability.

@@

  \subsection{LoRs from Admin People}\label{sec:admin-letters}

  Many international students try to get LoRs from high-ranking administrators in their universities such as department chair/head, dean, or director. The students never worked with these people (they might take a class or so with these profs), but mistakenly believe that these LoRs are valuable due to the writer's high position in the university.
  However, as emphasized, a generic LoR has little value because the writer does not know you well and can talk in depth about your research ability.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
Many students write letters themselves, typically due to the requests of their profs. or letter writers. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in it is you who write about yourself (and they just sign the letter).

Adcom members are just too experienced (having reviews hundreds of those letters) to spot and suspect such letters.  There is a very difference between a letter written by an experienced professor and a student.  For example, they can provide convincing and concrete examples based on their experience and compare you to their own students, and of course the writing style of a professor is very different from an undergraduate student.  Worse yet, if we suspect that the student wrote the letter, it is a \red{red flag} as we will question both the student's integrity and the letter writer's credibility.

@@

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  You can explicitly ask if they are willing to write a strong letter for you.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

@@

 As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

>>>>

\chapter{Letters of Recommendation (LoR)}\label{chap:lor}
\chapterinfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

Most PhD programs will require at least \emph{two} letter of recommendations (LoRs). 
The goal of the LoRs is to provide the adcom with an assessment of your research ability from someone who knows you.  The LoRs should focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher to adcom.

@@

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also \emph{have the credibility} to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD, so they know what a PhD is about and so can evaluate your research ability. LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\subsection{LoR from Famous People}  Having a strong letter from an internationally recognized researcher can \emph{greatly strengthen} your application. Such a letter can outweigh the lack of publications (or low GPA or unknown school). A well-known researcher has a reputation to maintain (so they won't recommend someone bad), and adcom member trust people that they know, or heard about and respect (compared to someone they don't know).

@@

%It is also worth noting that a LoR from a ``full'' senior professor who has not been active in research might not be as strong as a LoR from a junior faculty who has been actively mentoring you in research.  This also applies to administrative, e.g., chairs, directors, and deans (\autoref{sec:admin-letters}.) The fame of the writer is less important than how well they know you and can write about your research ability.

@@

  \subsection{LoRs from Dept Chair, Dean, or Boss and Supervisor at Work}\label{sec:admin-letters}

  Many students try to get LoRs from high-ranking administrators in their universities such as department chair/head, dean, or director. The students never worked with these people (they might take a class or so with these profs), but mistakenly believe that these LoRs are valuable due to the writer's high position in the university.
  However, as emphasized, a generic LoR has little value because the writer does not know you well and can talk in depth about your research ability.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}

Many profs. and letter writers ask students to write their own letters, which is a common practice in some countries. Unfortunately, such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in it is you who write about yourself (and they just sign the letter).

Self-written letters are easy to spot as an experienced professor would write very differently compared to an undergraduate student.
For example, they can provide convincing and concrete examples based on their experience and compare you to their own students, and of course the writing style of a professor is very different from an undergraduate student.  Worse yet, if we suspect that the student wrote the letter, it is a \red{red flag} as we will question both the student's integrity and the letter writer's credibility.

@@

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  In the US, students usually explicitly ask if the writer would be willing to provide a strong letter, and the writer are also very direct in their response.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

@@

 As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you can provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
